DOpElib:\\ Differential Equations and Optimization Environment

### 2013-12-09 Examples/Documentation/foreword.tex
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The aim of DOpE is to provide a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by PDE. The
solution of a broad variety of PDE is possible in deal.II as well, but
DOpE concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. While deal.II leaves much of the work and many
decisions to the user, DOpE intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the DOpE framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
At the present stage the following features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from deal.II including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Mesh adaptation.
\item Different spatial triangulations for control and state variables.
\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The aim of DOpE is to provide a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by PDE. The
solution of a broad variety of PDE is possible in deal.II as well, but
DOpE concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. 
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and multiphysics problems such as 
fluid-structure interactions.

\twick{This is copied from the paper:
The main feature of \dope{} is to give a unified interface to high level 
algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed in such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.
In addition to the finite element code provided by deal.II --which at present is the only FE-toolkit to which we provide an 
interface-- 
the presented library
\dope{} is user-focused by delivering
prefabricated tools which require adjustments by the user only for parts
connected to his specific problem. 
This is in contrast to deal.II which leaves the implementation of all high-level algorithms to the user.
The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.
}

\twickremove{Please check if we can remove part and use the red part!
While deal.II leaves much of the work and many
decisions to the user, DOpE intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the DOpE framework.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
At the present stage the following features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from deal.II including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, plasticity, fluid-structure interaction problems,
and finally a coupled Biot-Lam\'e-Navier system
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Mesh adaptation.
\item Different spatial triangulations for control and state variables.
\end{itemize}

### 2013-12-09 Examples/OPT/StatPDE/Example1/content.tex
commit before talk of t and w
Code cleaned up to OPT StatPDE Ex2. TODO: content.tex
contents tex in opt problems
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

\subsubsection{General problem description}
This example solves the distributed minimization problem
\begin{gather*}
\min J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) = (q+f,\phi)\;\;\forall\,\phi \in H^1_0(\Omega)
\end{gather*}
on the domain $\Omega = [0,1]^2$, and the data is chosen as follows:
\begin{gather*}
 f = \left(20\pi^2  \sin(4 \pi x) - \frac{1}{\alpha}  \sin(\pi x)\right) \sin(2 \pi y)\\
 u^d = \left( 5 \pi^2 \sin(\pi x) + \sin(4 \pi x)\right)  \sin(2\pi y)
\end{gather*}
and $\alpha = 10^{-3}$.
Hence its solution is given by:
\begin{gather*}
 \overline{q} = \frac{1}{\alpha} \sin(\pi x) \sin(2 \pi y)\\
 \overline{u} = \sin(4 \pi x) \sin(2 \pi y).
\end{gather*}
Thus the exact optimal value of the cost functional can be calculated as 
\begin{gather*}
 J^* = J(\overline{q},\overline{u}) = \frac{1}{8}\Bigl(25\pi^4 + \frac{1}{\alpha}\Bigr).
\end{gather*}

@@

The example shows how to estimate the error in the cost functional for stationary optimization
problems.

>>>>

\subsubsection{General problem description}
\twick{Dieses Beispiel ist nun extremst ausfuehrlich hinsichtlich der
localpde. Da dies ein Manual ist, wuensche ich mir, dass wir dies (respective
moeglicher Rechtschreibfehler und typos) auch so belassen. Mir hat das damals 
bei FSI Opt sehr geholfen.}
This example solves a distributed minimization problem
\twick{and shows how to estimate the error in the cost functional for stationary optimization
problems. The problem reads:}
\begin{gather*}
\min J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) = (q+f,\phi)\;\;\forall\,\phi \in H^1_0(\Omega)
\end{gather*}
on the domain $\Omega = [0,1]^2$, and the data is chosen as follows:
\begin{gather*}
 f = \left(20\pi^2  \sin(4 \pi x) - \frac{1}{\alpha}  \sin(\pi x)\right) \sin(2 \pi y)\\
 u^d = \left( 5 \pi^2 \sin(\pi x) + \sin(4 \pi x)\right)  \sin(2\pi y)
\end{gather*}
and $\alpha = 10^{-3}$.
Hence its solution is given by:
\begin{gather*}
 \overline{q} = \frac{1}{\alpha} \sin(\pi x) \sin(2 \pi y)\\
 \overline{u} = \sin(4 \pi x) \sin(2 \pi y).
\end{gather*}
Thus the exact optimal value of the cost functional can be calculated as 
\begin{gather*}
 J^* = J(\overline{q},\overline{u}) = \frac{1}{8}\Bigl(25\pi^4 + \frac{1}{\alpha}\Bigr).
\end{gather*}

@@

\twick{Ab hier ist alles neu!!}
\subsubsection{Background information and program description}
In the following, we describe all extensions to the previous problems 
relevant to solving PDE-based optimization with \dope{}. So far, 
we had only to implement the \texttt{ElementEquation} and the corresponding
matrix \texttt{ElementMatrix}. Now, based on the idea of the reduced cost
functional, we have to compute certain additional equations representing the 
adjoint, tangent, and adjoint hessian equations
\texttt{ElementEquation\underline{ }U, ElementEquation\underline{ }T,
ElementEquation\underline{ }UTT}
for the state equation and in the same terms arising from the functional
itself. Let us shed some light into all equations by giving some background
information and overview first.

{
In abstract form, we are given the following optimization problem:
\begin{equation*}
J(q,U) \, \rightarrow \, \min , \quad a(q,U)(\Psi) = 0 \quad \forall\Psi V
\end{equation*}
Lagrangian:
\begin{equation*}
{\cal L}(q,U,Z):=J(q,U) - a(q,U)(Z)
\end{equation*}
Optimality system (KKT system):
\begin{align*}
a_U'(q,U)(\Phi , Z) &= J'_U(q,U)(\Phi) \quad \forall \Phi\in V \\
a_q'(q,U)(\chi , Z) &= J'_q(q,U)(\chi) \quad \forall \chi\in Q \\
a(q,U)(\Psi) &= 0 \qquad \forall\Psi\in V
\end{align*}
or equivalently
\begin{align*}
{\cal L}'_U(q,U,Z)(\Phi) &= 0 \quad\forall \Phi\in V \quad\text{(Adjoint Equation)} \\
{\cal L}'_q(q,U,Z)(\chi) &= 0 \quad\forall \chi\in V \quad\text{(Gradient Equation)} \\
{\cal L}'_Z(q,U,Z)(\Psi) &= 0 \quad\forall \Psi\in V \quad\text{(State Equation)}
\end{align*}
The continuous problem is discretized by a standard Galerkin method using 
finite dimensional subspaces $Q_h \times V_h \subset Q\times V$:
\begin{equation*}
J(q,U) \, \rightarrow \, \min , \quad a(q,U)(\Psi) = 0 \quad \forall\Psi V
\end{equation*}
Discrete saddle-point problems
\begin{align*}
a_U'(q_h,U_h)(\Phi_h , Z_h) &= J'_U(q_h,U_h)(\Phi_h) \quad \forall \Phi_h\in V_h \\
a_q'(q_h,U_h)(\chi_h , Z_h) &= J'_q(q_h,U_h)(\chi_h) \quad \forall \chi_h\in Q_h \\
a(q_h,U_h)(\Psi_h) &= 0 \qquad \forall\Psi_h\in V_h
\end{align*}
\subsubsection{Solution process}
In this section, we briefly discuss the solution process for the 
optimization problem. We further details, we refer to the standard literature.
The unconstrained optimal control problem is reformulated as follows. 
We introduce the solution operator $S:Q\rightarrow V$ of the state equation.
Then:
\begin{equation*}
j(q):= J(q,S(q))\, \rightarrow \, \min , \quad a(q,S(q))(\Psi) = 0 \quad 
\forall \Psi \in V.
\end{equation*}
The local existence and sufficient regularity of $S$ is assumed. The 
necessary optimality conditions of first and second order are
\begin{equation*}
j'(q)(\delta q) = 0 , \quad j''(q)(\delta q,\delta q) \geq 0 \quad 
\forall\delta q\in Q.
\end{equation*} 
The derivatives of the reduced functional can be computed using the 
Lagrangian
\begin{equation*}
{\cal L}(q,U,Z) =J(q,U) - a(q,U)(Z)
\end{equation*}
as already introduced. Let $q\in Q$ be given. We derive
 the corresponding state $U=S(q)\in V$ and the dual variable
$Z\in V$. The following terms hold:
}

\textbf{Dual equation}

\begin{equation*}
{\cal L}_U'(q,U,Z)(\Psi) = 0 \quad\forall\Psi\in V
\end{equation*}
Then,
\begin{equation*}
j'(q)(\delta q) = {\cal L}_q' (q,U,Z)(\delta q) \quad \text{for }\delta q\in Q.
\end{equation*}

\textbf{Tangent equation}

For the right-hand side of Newton's method, we also need further terms. Let
$\delta q\in Q$ be given, and let $\delta U\in V$ and $\delta Z\in V$ fulfill
the equation
\begin{equation*}
{\cal L}_{qZ}'' (q,U,Z)(\delta q,\Phi) + {\cal L}_{UZ}'' (q,U,Z)(\delta
U,\Phi)
\quad\forall \Phi\in V,
\end{equation*}
further

\textbf{Dual Hessian equation}

\begin{equation*}
{\cal L}_{qU}'' (q,U,Z)(\delta q,\Phi) + {\cal L}_{UU}'' 
(q,U,Z)(\delta U,\Phi) + {\cal L}_{ZU}'' (q,U,Z)(\delta Z,\Phi)
\quad\forall \Phi\in V,
\end{equation*}
Then, for $\delta r\in Q$, it holds
\begin{align*}
j''(q)(\delta q, \delta r) 
&= {\cal L}_{qq}'' (q,U,Z)(\delta q,\delta r) \\
&+ {\cal L}_{Uq}'' (q,U,Z)(\delta U,\delta r) \\ 
&+ {\cal L}_{Zq}'' (q,U,Z)(\delta Z,\delta r) \quad \text{for }\delta r\in Q.
\end{align*}
These terms build the the Newton scheme as follows:
\begin{align*}
j''(q^n )(\delta q,\chi ) =&\, -j'(q^n)(\chi) \quad\forall\chi \in Q,\\
q^{n+1} :=&\, q^n + \delta q , \quad n=0,1,2,\ldots .
\end{align*}

Finally, we collect the representations of the forgoing work. First, we recall 
the terms of the Lagrangian:

\textbf{Adjoint:} Find $z\in X$ such that
\begin{equation*}
{\cal L}_u' (q,u,z)(\phi) = 0 \quad \forall\phi\in X. 
\end{equation*}
\textbf{Tangent (linearized state):} Find $\delta u\in X$ such that
\begin{equation*}
{\cal L}_{qz}'' (q,u,z)(\delta q, \phi) +
{\cal L}_{uz}'' (q,u,z)(\delta u, \phi) = 0 \quad \forall\phi\in X. 
\end{equation*}
\textbf{Additional adjoint:} Find $\delta z\in X$ such that
\begin{equation*}
{\cal L}_{qu}'' (q,u,z)(\delta q, \phi) +
{\cal L}_{uu}'' (q,u,z)(\delta u, \phi) +
{\cal L}_{zu}'' (q,u,z)(\delta z, \phi) = 0 \quad \forall\phi\in X. 
\end{equation*}

Especially, the equations for stationary optmization problems are
given as follows:

\textbf{Adjoint:} Find $z\in X$ such that
\begin{equation*}
a_u'(q,u)(\phi,z) = J'(u)(\phi) \quad \forall\phi\in X. 
\end{equation*}
\textbf{Tangent:} Find $\delta u\in X$ such that
\begin{equation*}
a_u'(q,u)(\delta u,\phi) = -a_q'(q,u)(\delta q,\phi)  \quad \forall\phi\in X. 
\end{equation*}
\textbf{Additional adjoint:} Find $\delta z\in X$ such that
\begin{equation*}
a_u'(q,u)(\phi,\delta z) = a_{uu}''(q,u)(\delta u,\phi,z)
-a_{qu}''(q,u)(\delta q,\phi,z) + J''(u)(\delta,\phi) \quad \forall\phi\in X. 
\end{equation*}

Using the Lagrangian, the previous expressions read

\textbf{Lagrangian}
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}

\textbf{State Equation}
\begin{align*}
a(q,u)(\phi) = 0 
\end{align*}

\textbf{Dual Equation}

\begin{align*}
{\cal L}_{u}' (q,u,z)(\phi ) &= 0 \\
\Leftrightarrow a_u' (q,u)(\phi,z) &= J_u' (q,u)(\phi)
\end{align*}

\textbf{Tangent Equation}

\begin{align*}
{\cal L}_{qz}'' (q,u,z)(\delta q, \phi) 
+ {\cal L}_{uz}'' (q,u,z)(\delta u, \phi) &= 0 \\
\Leftrightarrow a_u' (q,u)(\delta u,\phi) + a_q' (q,u)(\delta q,\phi) &= 0
\end{align*}

\textbf{Dual Hessian Equation}

\begin{align*}
{\cal L}_{qu}'' (q,u,z)(\delta q, \phi) 
&+ {\cal L}_{uu}'' (q,u,z)(\delta u, \phi) \\
&+ {\cal L}_{zu}'' (q,u,z)(\delta z, \phi) = 0 \\
\Leftrightarrow a_u' (q,u)(\phi, \delta z) 
&+ a_{uu}' (q,u)(\delta u,\phi, z) \\
&+ a_{qu}' (q,u)(\delta q,\phi, z)  = J_{uu}''(q,u)(\cdot ,\cdot )
\end{align*}

\textbf{Newton left hand side: first derivative}

\begin{align*}
j'(q)(\delta q) &= {\cal L}_{q}' (q,u,z)(\delta q) \quad\text{(Gradient equation)} \\
\Leftrightarrow j'(q)(\delta q) &= -a_q' (q,u)(\delta q,z) + J_{q}'(q,u)(\delta q)
\end{align*}

\textbf{Newton right hand side: second derivative}

\begin{align*}
j''(q)(\delta q, \delta r) &= {\cal L}_{qq}'' (q,u,z)(\delta q, \delta r) + 
{\cal L}_{uq}'' (q,u,z)(\delta u, \delta r) \\
&\quad +  {\cal L}_{zq}'' (q,u,z)(\delta z, \delta r) \\
\Leftrightarrow  j''(q)(\delta q, \delta r) &= 
J_{qq}''(q,u)(\delta q, \delta r) -a_{qq}'' (q,u)(\delta q,\delta r, z) \\
&\quad - a_{uq}'' (q,u)(\delta u,\delta r, z) -a_{q}' (q,u)(\delta r,\delta z)  
\end{align*}

\subsubsection{Implementation in \dope{}}

\begin{align*}
\text{ElementEquation (state)} \quad &\Leftrightarrow \quad a(q,u)(\phi) \\
\text{ElementEquationU (adjoint)} \quad 
&\Leftrightarrow \quad a_u'(q,u)(\phi, z) \\ 
\text{ElementEquationUT (tangent)} \quad 
&\Leftrightarrow \quad a_u'(q,u)(\delta u, \phi) \\
\text{ElementEquationUTT (adjoint hessian)} \quad 
&\Leftrightarrow \quad  a_u'(q,u)(\phi, \delta z) \\
\text{ElementEquationQ (gradient)} \quad 
&\Leftrightarrow \quad a_q'(q,u)(\delta q,z) \\
\text{ElementEquationQT (tangent)} \quad 
&\Leftrightarrow \quad a_q'(q,u)(\delta q, \phi) \\
\text{ElementEquationQTT (hessian)} \quad 
&\Leftrightarrow \quad a_{q}'(q,u)(\delta q,\delta z) \\
\text{ElementEquationUU (adjoint hessian)} \quad 
&\Leftrightarrow \quad a_{uu}''(q,u)(\delta u, \phi, z) \\
\text{ElementEquationQU (adjoint hessian)} \quad 
&\Leftrightarrow \quad a_{qu}''(q,u)(\delta q, \phi, z) \\
\text{ElementEquationUQ (hessian)} \quad 
&\Leftrightarrow \quad a_{uq}''(q,u)(\delta u, \delta r, z)\\
\text{ElementEquationQQ (hessian)} \quad 
&\Leftrightarrow \quad a_{qq}''(q,u)(\delta q, \delta r, z)\\
\text{ElementRightHandSide (state)} \quad 
&\Leftrightarrow \quad f(\phi ) \\
\text{ElementMatrix (all)} \quad 
&\Leftrightarrow \quad a'(q,u)(\delta u) \\
\text{ControlElementEquation (gradient or hessian)} \quad 
&\Leftrightarrow \quad {\cal L}_q' (q,u,z)(\delta q) \\
\text{ControlElementMatrix (all)} \quad 
&\Leftrightarrow \quad a(q,u)(\phi)
\end{align*}

Functionals
\begin{align*}
\text{ElementValue (all)} \quad  & \Leftrightarrow \quad J(q,u) \\
\text{ElementValueU (all)} \quad  & \Leftrightarrow \quad J_u'(q,u)(\phi) \\
\text{ElementValueQ (all)} \quad  & \Leftrightarrow \quad J_q'(q,u)(\phi) \\
\text{ElementValueUU (all)} \quad & \Leftrightarrow \quad J_{uu}''(q,u)(\psi, \phi) \\
\text{ElementValueQU (all)} \quad & \Leftrightarrow \quad J_{qu}''(q,u)(\psi, \phi) \\
\text{ElementValueUQ (all)} \quad & \Leftrightarrow \quad J_{uq}''(q,u)(\psi, \phi) \\
\text{ElementValueQQ (all)} \quad & \Leftrightarrow \quad J_{qq}''(q,u)(\psi, \phi) \\
\end{align*}

\subsubsection{Back to the specific equations in this example}
We have
\[
a(q,u)(\phi) = (\nabla u, \nabla\phi) - (q + f, \phi).
\]
and
\[
a_u'(q,u)(\phi, z) = (\nabla\phi, \nabla z)
\]
and
\[
a_u'(q,u)(\delta u, \phi) = (\nabla\delta u, \nabla\phi)
\]
and
\[
a_u'(q,u)(\phi, \delta z) = (\nabla\phi, \nabla\delta z).
\]
and 
\[
a_q'(q,u)(\delta q, \phi) = -(z,\psi^q)
\]
and 
\[
a_q'(q,u)(\delta q, \phi) = (\delta q,\phi)
\]
and finally
\[
a_{q}'(q,u)(\delta q,\delta z) = - (\delta z,\psi^q)
\]
On the functional level, we have the following terms:
\[
J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2d
\]
and 
\[
J_u'(q,u)(\phi) = (u-u^d, \phi)
\]
and
\[
J_q'(q,u)(\phi) = \alpha (q,\psi^q)
\]
and 
\[
J_{uu}''(q,u)(\psi, \phi) = (\delta u, \phi).
\]
All other terms, specifically mixed terms with $QU$ etc. are zero in this
example.

\subsubsection{\texttt{main.cc}}
Finally, the main file of the optimization examples does not look very much 
different than for pure PDE computations - which is one of the crucial aims 
of our library. 
Here, instead of using a \texttt{pdeproblemcontainer}, we use 
now an \texttt{optproblemcontainer}. Furthermore, 
we define \texttt{ReducedNewtonAlgorithm} to solve the optimization problem.
Next, in the body of the main file, we introduce a second FE function for the 
control variable (using the FE\underline{}Nothing function as we will see 
in Example 2 also allows for 0d controls). Then, we define 
a \texttt{COSTFUNCTIONAL}. Finally, the problem is either solved 
by calling \texttt{Alg.Solve(q)} and/or the user might check if the 
derivatives are implemented correctly by calling 
\texttt{Alg.CheckGrads} or \texttt{Alg.CheckHessian}. The latter 
two functionalities are highly recommended to check your initial 
implementation before wondering about your results.

### 2013-12-09 Examples/OPT/StatPDE/Example2/content.tex
Code cleaned up to OPT StatPDE Ex2. TODO: content.tex
Further changes
Overhaul of OPT StatPDe Example 5
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) &= (f(q),\phi)\;\;\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
Notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. So our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement 
\begin{itemize} 
\item \texttt{Value},
\item  \texttt{Value\_U},
\item \texttt{Value\_Q}, 
\item \texttt{Value\_UU},
\item \texttt{Value\_UQ},
\item \texttt{Value\_QU},
\item \texttt{Value\_QQ}
\end{itemize}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceeding \text{Point} (\texttt{PointValue} etc.).

>>>>

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2&&\\
\text{s.t. } (\nabla u,\nabla \phi) &= (f(q),\phi) \quad\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with zero dirichlet boundary conditions and
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
\twick{
In contrast to the first example, the control is now a discrete quantity (for
the three observation points) where we use the \texttt{FE\underline{}Nothing}
element to assigne the three controls. Here, the number of components equals
the number of controls. 
In addition, }notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. To handle this, we need a special integrator as well as a special newtonsolver. Additionally, our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement 
\begin{itemize} 
\item \texttt{ElementValue},
\item \texttt{ElementValue\_U},
\item \texttt{ElementValue\_Q}, 
\item \texttt{ElementValue\_UU},
\item \texttt{ElementValue\_UQ},
\item \texttt{ElementValue\_QU},
\item \texttt{ElementValue\_QQ}
\end{itemize}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceeding \text{Point} (\texttt{PointValue} etc.).
%FE_Nothing, handle of the control erlaeutern.

### 2013-12-09 Examples/OPT/StatPDE/Example3/content.tex
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

The target functional is considered as 
\[
k(v,p) = \int_{\Gamma_O} n\cdot \sigma(v,p)\cdot d \, ds,
\]
where $\Gamma_O$ denotes the cylinder boundary, and $d$ is a vector in the
direction
of the mean flow. For theoretical and numerical reasons, this functional 
needs to be regularized, including the control variable $q$, such that
\[
K(v,p,q) = k(v,p) + \frac{\alpha}{2}||q - q_0||,
\] 
where $\alpha$ is the Tikhonov parameter and $q_0$ some 
reference control.

The rest of the progam is similar to the previous optimization problems where
we formulate the state equation in a weak form $a(v,p)(\phi)$ such that the 
final problem reads
\[
K(v,p,q) \rightarrow 0 \quad \text{s.t.} \quad a(v,p)(\phi) = 0.
\]

%\subsubsection{Program description}

>>>>

The control $q$ enters via the weak formulation. It reads,
\begin{align*}
a(q,v,p)(\phi) = (\sigma(v,p), \phi) + (v\cdot \nabla v, \phi) 
+ \langle q, \phi\cdot n \rangle + (\nabla\cdot v, \chi) = 0,
\end{align*}

The target functional is considered as 
\[
k(v,p) = \int_{\Gamma_O} n\cdot \sigma(v,p)\cdot d \, ds,
\]
where $\Gamma_O$ denotes the cylinder boundary, and $d$ is a vector in the
direction
of the mean flow. For theoretical and numerical reasons, this functional 
needs to be regularized, including the control variable $q$, such that
\[
K(q,v,p) = k(v,p) + \frac{\alpha}{2}||q - q_0||,
\] 
where $\alpha$ is the Tikhonov parameter and $q_0$ some 
reference control.

The rest of the progam is similar to the previous optimization problems where
we formulate the state equation in a weak form $a(v,p)(\phi)$ such that the 
final problem reads
\[
K(q,v,p) \rightarrow \text{min} \quad \text{s.t.} \quad a(q,v,p)(\phi) = 0.
\]

\twick{
\subsubsection{Program description}
The implementation of this example does not introduce any new \dope{}-specific
features but shows that more complicated equations such as the Navier-Stokes 
system can be used as forward problem for the optimization process. 
This example builds on the previous Example \ref{OPT_Stat_Param_Lin_Ellipt}.
}

### 2013-12-09 Examples/OPT/StatPDE/Example7/content.tex
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

The Problem is similar to that of {\tt OPT/StatPDE/Example1} except for the box control constraints.
The problem is solved using the optimization library IPOPT that you can obtain for free. To use it 
a correct link to the ipopt library needs to be created in  {\tt DOpE/ThirdPartyLibs} by the name 
{\tt ipopt}, i.e., you should have the file
{\tt DOpE/ThirdPartyLibs/ipopt} pointing to the ipopt directory. If you have not done this you can compile the
example but when running the example you will only get an error message like\\
{\tt Warning: During execution of `Reduced\underline{ }IpoptAlgorithm::Solve`\\ the following Problem occurred!\\
To use this algorithm you need to have IPOPT installed! \\To use this set the WITH\underline{ }IPOPT CompilerFlag.}

>>>>

\subsubsection{Program description}
\twick{
The Problem is similar to that of {\tt OPT/StatPDE/Example1} except for the box control constraints.
The implementation of these constraints is taken care of in the main file
where we add a constrained description \texttt{lcc}.
}

\twickremove{Winni: in dieser Beischreibung ist 
\[
-500 \leq q \leq 500.
\]
Allerdings finde ich die '500' nirgendwo in main, localpde, or localfunctional
file. Wie funktioniert dieses Programm'?}

\subsubsection{External optimization solver}
The problem is solved using the optimization library IPOPT that you can obtain for free. To use it 
a correct link to the ipopt library needs to be created in  {\tt DOpE/ThirdPartyLibs} by the name 
{\tt ipopt}, i.e., you should have the file
{\tt DOpE/ThirdPartyLibs/ipopt} pointing to the ipopt directory. If you have not done this you can compile the
example but when running the example you will only get an error message like\\
{\tt Warning: During execution of `Reduced\underline{ }IpoptAlgorithm::Solve`\\ the following Problem occurred!\\
To use this algorithm you need to have IPOPT installed! \\To use this set the WITH\underline{ }IPOPT CompilerFlag.}

### 2013-12-09 Examples/OPT/StatPDE/Example8/content.tex
1) Minor changes in the formatting of the documentation. 2) Added documentation to PDE/StatPDE/Example9
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

This example implements the topology optimization of an MBB-Beam given in \\
{\tt OPT/StatPDE/Example6} using the SIMP method.

@@

>>>>

\subsubsection{General problem description}
This example implements the topology optimization of an MBB-Beam given in \\
{\tt OPT/StatPDE/Example6} using the SIMP method.

@@

\subsubsection{Program description}
\twick{
The global constraint is inserted in the main file via 
\texttt{constraints(lcc, 1)} instead of 
\texttt{constraints(lcc, 0)} what we had in the previous two examples.
}

### 2013-12-09 Examples/OPT/StatPDE/Example9/content.tex
Update Doku und Beseitigung kleinerer Doku-typos in zwei main files.

>>>>

\subsubsection{Program description}
\twick{
This example is similar to Example \ref{OPT_Stat_Param_Nonlin_Fluid} (both 
based on Example \ref{OPT_Stat_Param_Lin_Ellipt}), except that we 
have again much more complicated (nonlinear) equations. A modification 
of this example and several numerical tests are presented 
in \cite{RiWi13_fsi_opt}.
}
